DATA 621---Business Analytics and Data Mining

### 2020-09-09 HW1/HW1_AF.Rmd
data exploration updates

## 1. DATA EXPLORATION (Amber)
Describe the size and the variables in the moneyball training data set. Consider that too much detail will cause a manager to lose interest while too little detail will make the manager consider that you aren’t doing your job. Some suggestions are given below. Please do NOT treat this as a check list of things to do to complete the assignment. You should have your own thoughts on what to tell the boss. These are just ideas.

a. Mean / Standard Deviation / Median
b. Bar Chart or Box Plot of the data
c. Is the data correlated to the target variable (or to other variables?)
d. Are any of the variables missing and need to be imputed “fixed”?

## 2. DATA PREPARATION (25 Points)
Describe how you have transformed the data by changing the original variables or creating new variables. If you did transform the data or create new variables, discuss why you did this. Here are some possible transformations.

a. Fix missing values (maybe with a Mean or Median value)
b. Create flags to suggest if a variable was missing
c. Transform data by putting it into buckets
d. Mathematical transforms such as log or square root (or use Box-Cox)
e. Combine variables (such as ratios or adding or multiplying) to create new variables

## 3. BUILD MODELS (25 Points)
Using the training data set, build at least three different multiple linear regression models, using different variables (or the same variables with different transformations). Since we have not yet covered automated variable selection methods, you should select the variables manually (unless you previously learned Forward or Stepwise selection, etc.). Since you manually selected a variable for inclusion into the model or exclusion into the model, indicate why this was done.

Discuss the coefficients in the models, do they make sense? For example, if a team hits a lot of Home Runs, it would be reasonably expected that such a team would win more games. However, if the coefficient is negative (suggesting that the team would lose more games), then that needs to be discussed. Are you keeping the model even though it is counter intuitive? Why? The boss needs to know.

## 4. SELECT MODELS (25 Points)
Decide on the criteria for selecting the best multiple linear regression model. Will you select a model with slightly worse performance if it makes more sense or is more parsimonious? Discuss why you selected your model.

For the multiple linear regression model, will you use a metric such as Adjusted R2, RMSE, etc.? Be sure to explain how you can make inferences from the model, discuss multi-collinearity issues (if any), and discuss other relevant model output. Using the training data set, evaluate the multiple linear regression model based on (a) mean squared error, (b) R2, (c) F-statistic, and (d) residual plots. Make predictions using the evaluation data set.

## APPENDIX
### Assignment Overview
In this homework assignment, you will explore, analyze and model a data set containing approximately 2200
records. Each record represents a professional baseball team from the years 1871 to 2006 inclusive. Each record
has the performance of the team for the given year, with all of the statistics adjusted to match the performance of
a 162 game season.

Your objective is to build a multiple linear regression model on the training data to predict the number of wins
for the team. You can only use the variables given to you (or variables that you derive from the variables
provided). Below is a short description of the variables of interest in the data set:

\begin{center}
\begin{tabular}{|l|l|l|}
\hline
\textbf{VARIABLE NAME} & \textbf{DEFINITION} & \textbf{THEORETICAL EFFECT} \\ \hline
INDEX & Identification Variable (do not use) & None \\ \hline
TARGET\_WINS & Number of wins & \\ \hline
TEAM\_BATTING\_H & Base Hits by batters (1B,2B,3B,HR) & Positive Impact on Wins  \\ \hline
TEAM\_BATTING\_2B & Doubles by batters (2B) & Positive Impact on Wins \\ \hline
TEAM\_BATTING\_3B & Triples by batters (3B) & Positive Impact on Wins \\ \hline
TEAM\_BATTING\_HR & Homeruns by batters (4B) & Positive Impact on Wins \\ \hline
TEAM\_BATTING\_BB & Walks by batters & Positive Impact on Wins \\ \hline
TEAM\_BATTING\_HBP & Batters hit by pitch (get a free base) & Positive Impact on Wins \\ \hline
TEAM\_BATTING\_SO & Strikeouts by batters & Negative Impact on Wins \\ \hline
TEAM\_BASERUN\_SB & Stolen bases & Positive Impact on Wins \\ \hline
TEAM\_BASERUN\_CS & Caught stealing & Negative Impact on Wins \\ \hline
TEAM\_FIELDING\_E & Errors & Negative Impact on Wins \\ \hline
TEAM\_FIELDING\_DP & Double Plays & Positive Impact on Wins \\ \hline
TEAM\_PITCHING\_BB & Walks allowed & Negative Impact on Wins \\ \hline
TEAM\_PITCHING\_H & Hits allowed & Negative Impact on Wins \\ \hline
TEAM\_PITCHING\_HR & Homeruns allowed & Negative Impact on Wins \\ \hline
TEAM\_PITCHING\_SO & Strikeouts by pitchers & Positive Impact on Wins \\ \hline

\end{tabular}
\end{center}

### R Programming code

```{r}

train_set <- read.csv('data/moneyball-training-data.csv')
eval_set <- read.csv('data/moneyball-evaluation-data.csv')

colnames(train_set)

@@

>>>>

```{r loadData, include=FALSE}
library(tidyverse)
library(ggplot2)
library(corrplot)
train_set <- read.csv('data/moneyball-training-data.csv')
eval_set <- read.csv('data/moneyball-evaluation-data.csv')
```

## 1. DATA EXPLORATION

### How often does the team win?

We are given a data set containing 2,276 records, each of which includes the 
number of wins the team had that year and 15 additional statistics about the 
season. The team wins on average about 50% of the games they play (81 games 
out of 162), with the best season having 146 wins and the worst season having 
0 wins. The data is pretty normally distributed, with most seasons having between
49 and 112 wins (blue lines below). This means that there aren't too many extreme 
seasons where the team won significantly more or significantly less games than 
usual. This serves as a good gut-check for our final predictions; if the 
predicted wins are too high or too low, we know something is probably off.

```{r, echo = FALSE, message = FALSE, warning = FALSE, fig.height = 3, fig.width = 5, fig.align = 'center'}

avgWins <- mean(train_set$TARGET_WINS)
sdev <- sd(train_set$TARGET_WINS)
l1 <- mean(train_set$TARGET_WINS) - (2*sdev)
l2 <- mean(train_set$TARGET_WINS) + (2*sdev)

# histogram of wins
p <- ggplot(train_set, aes(x=TARGET_WINS)) + 
  geom_histogram() + 
  geom_vline(aes(xintercept= avgWins),
            color="red", linetype="dashed", size=1)+
  geom_vline(aes(xintercept=l1),
            color="blue", linetype="dashed", size=1) +
    geom_vline(aes(xintercept=l2),
            color="blue", linetype="dashed", size=1) +
  labs(title="Distribution of Wins",x="Number of wins", y = "Number of seasons")

p
```

### What's missing? 
A first look at the data shows that only about 8% of the records have 
a full set of information. The good news is that most of the missing values come 
from what we would expect because they don't happen too often; hit-by-pitch 
(92% missing!), caught stealing (34% missing), and double plays (13% missing). 
Since we have so little hit-by-pitch data, we expect that it doesn't contribute 
to overall wins and will eliminate it from a few of the models we propose. The other
two stats have less than half of the data missing, so we'll need to think of a 
clever way to impute these values. The remaining missing information is from a 
combination of stolen bases and strikeouts (by batters and pitchers). **It seems** 
**completely unreasonable** to have zero strike outs in a season, so this is something 
we'll most certainly have to impute.

```{r, include = FALSE}

pct_null <- data.frame(do.call("rbind", map(train_set %>% select(-INDEX), ~mean(is.na(.)))))
colnames(pct_null) <- c('PCT_NULL')

pct_null %>%
  arrange(desc(PCT_NULL)) %>%
  filter(PCT_NULL > 0)

@@

### Do the individual stats affect winning?
Intuitively, we expect that Caught stealing, Errors, Hits allowed, Homeruns allowed, 
Strikeouts by batters, and Walks allowed would all have a **negative** impact on
the total wins. In other words, as these stats increase, we would expect that the 
team is less likely to win.

However, when we look at the data for each of these statistics with respect to the 
total number of wins, we don't see a clear relationship across the board. In fact, 
only Errors and Hits allowed seem to have a negative impact on the wins. Caught stealing
and strikeouts by batters appear to be random -- whether the stat for a particular season
is high or low doesn't affect the overall number of wins.

Even more interestingly, Homeruns and Walks allowed have the *opposite* effect; as the number
of homeruns and walks that are allowed from the opposing team increase, the number of wins 
appears to increase as well!

```{r, echo = FALSE, message = FALSE, warning = FALSE}

negSet <- train_set %>%
  select(TARGET_WINS, 
         'Errors' = TEAM_FIELDING_E, 
         'Hits allowed' = TEAM_PITCHING_H, 
         'Strikeouts by batters' = TEAM_BATTING_SO, 
         'Caught stealing' = TEAM_BASERUN_CS, 
         'Walks allowed' = TEAM_PITCHING_BB, 
         'Homeruns allowed' = TEAM_PITCHING_HR)

ggplot(data = negSet %>%
  gather(-TARGET_WINS, key = "STAT", value = "VALUE"), aes(x = VALUE, y = TARGET_WINS)) + 
  geom_point() +
  geom_smooth(method = "lm", se = FALSE) +
  labs(title="Number of wins with respect to recorded stats",x="", y = "Number of Wins") +
  facet_wrap(~ STAT, scales = "free")

We can look at the same information for the stats that we expect to have a **positive**
effect on wins: Base hits, Doubles, Triples, Homeruns, and Walks by batters, Batters
getting hit by pitches, Stolen bases, Double Plays, and Strikeouts by pitchers.

Many of these stats *do* seem to have an effect on the number of wins. Most noteably,
as the number of base hits and walks increase, the team is more likely to win. We see
a slighter, positive relationship between homeruns, doubles, triples, and stolen bases.
This makes sense when we think about it -- these things tend to happen less often in
games than pure base hits and walks. Double plays and batters hit don't appear to have
an effect on the number of wins either. Once again, this intuitively makes sense because
they are less likely to happen in a game.

One thing to note is the number of strikeouts compared to the number of wins. We can see
that there are a few outliers -- abnormally high numbers of strikeouts in a season. This
should be taken with caution, as they don't represent a typical season's stats.

```{r, echo = FALSE, message = FALSE, warning = FALSE}

posSet <- train_set %>%
  select(TARGET_WINS, 
         'Base hits' = TEAM_BATTING_H, 
         'Doubles' = TEAM_BATTING_2B, 
         'Triples' = TEAM_BATTING_3B, 
         'Homeruns' = TEAM_BATTING_HR, 
         'Walks' = TEAM_BATTING_BB, 
         'Batters hit' = TEAM_BATTING_HBP,
         'Stolen bases' = TEAM_BASERUN_SB,
         'Double plays' = TEAM_FIELDING_DP,
         'Strikeouts' = TEAM_PITCHING_SO)

ggplot(data = posSet %>%
  gather(-TARGET_WINS, key = "STAT", value = "VALUE"), aes(x = VALUE, y = TARGET_WINS)) + 
  geom_point() +
  geom_smooth(method = "lm", se = FALSE) +
  labs(title="Number of wins with respect to recorded stats",x="", y = "Number of Wins") +
  facet_wrap(~ STAT, scales = "free")

Some of the provided statistics are well-balanced in the sense that there are
very few (or no) extreme values. **Homeruns by batters**, **Strikeouts by** 
**batters**, and **Homeruns allowed** are examples of this. Aside from

```{r, echo = FALSE, message = FALSE, warning = FALSE}

set1 <-  train_set %>%
  select('Homeruns by batters' = TEAM_BATTING_HR, 
         'Strikeouts by batters' = TEAM_BATTING_SO,  
         'Homeruns allowed' = TEAM_PITCHING_HR) %>%
  gather("stat", "value") %>%
  filter(complete.cases(.) == TRUE)

vals <- ggplot(set1 , aes(x=stat, y=value)) + 
    geom_boxplot() +
  labs(title="",x="", y = "Value") +
  facet_wrap(~stat, scale="free")

vals

# sapply(train_set %>% select(-INDEX, -TARGET_WINS), function(x){length(boxplot.stats(x)$out)})

```{r, echo = FALSE, message = FALSE, warning = FALSE}

# Correlation matrix with significance levels (p-value)
library("Hmisc")
res2 <- rcorr(as.matrix(train_set %>% select(-INDEX)))

# Insignificant correlation are crossed
corrplot(res2$r, type="upper", order="hclust", 
          tl.col = "black", p.mat = res2$P, sig.level = 0.01, insig = "blank")
